Generality in food webs scales with species richness but not with latitude

### 2015-06-10 manuscript/latitudepaper.tex
Updates from travelling
fixed merge remotely
All DBS comments are now incorporated

\title{Generality in food webs scales with species richness not not with latitude}
\author{Alyssa R. Cirtwill$^{1,2}$, Daniel B. Stouffer$^{1}$, Tamara N. Romanuk$^{2}$}
\date{$^1$School of Biological Sciences\\University of Canterbury\\
Private Bag 4800\\Christchurch 8140, New Zealand}
\medskip$^2$Department of Biology\\Life Science Centre\\
Dalhousie University\\1355 Oxford St.\\P0 BOX 15000\\
Halifax, NS B3H 4R2\\Canada\\}

@@

Several properties of food webs---the networks of feeding links between
species---vary systematically with the species richness of the underlying
community.  Under the ``latitude-niche breadth hypothesis'', which predicts
that species in the tropics will tend to evolve narrower niches, one might
expect that these scaling relationships could also be affected by latitude. To
test this hypothesis, we analysed the scaling relationships between species
richness and average generality, vulnerability, and links per species across a
set of 163 empirical food webs.  We also investigated scaling relationships
between the three food-web properties and the proportions of the web made up 
by basal resources, intermediate consumers, and top predators. While we
observed no effect of latitude on scaling relationships in the estuarine,
marine, and terrestrial food webs, there were strong effects of latitude on
scaling relationships in the freshwater food webs. In these communities,
the latitude-niche breadth hypothesis appears to hold true while in other habitat
types niches appear to be broader in the tropics. These contrasting findings
indicate that it is important to account for habitat type when exploring gradients
in food-web structure.

@@

  % [[list in supplemental (347 GlobalWeb webs \citep{GlobalWeb}), supplemented 
  % with 31 food webs used to study the relationship between consumer and resource body sizes (SizeWebs \citep{Brose2006})
  % and seven food webs used to study the impact of parasites on network structure (ParWebs \citep{Dunne2013}).
  % Of the GlobalWeb webs, 122 were rejected because their original source could not be found or was unpublished (53), they were source or sink
  % webs rather than descriptions of an entire community (13), were focused on plant-pollinator, host-parasitoid, or 
  % competitive interactions (8, 34, and 2, respectively), described inferred interactions in an extinct community (2), were ``generalized 
  % schemes'' rather than being based on empirical observation (8), or because it was not clear which of a variety of 
  % described sites the published food web represented (2). (See Table S1 for a list of rejected webs and reasons for their
  % exclusion.) For three GlobalWeb webs which included parasites as a minor component of the community, as well as the 
  % seven ParWebs webs, we included a modified version of each web which included only free-living species and the 
  % interactions between them \citep{Dunne2013}. 
  % As we were considering the log of proportions of basal resources, top predators, and intermediate consumers, each web used in our analysis 
  % had to have at least one species in each trophic group.

  % of  the University of Canberra's GlobalWeb database (www.globalwebdb.com) 
  % \citep{}. Of these, 302 webs had accessible original sources. 
  % 18 source and sink webs, which describe links to or from a focal taxon, were excluded as they are likely 
  % to have different food web properties than webs attempting to
  % describe the entire community of a site \citep{Williams2002}, as were 8 plant-pollinator, 34 host-parasitoid, and 2 
  % competition-focused networks
  % \citep{Riede2010}. Of the remaining 230 food webs, we excluded two paleowebs (comprised of probable interactions 
  % between extinct or prehistoric species) \citep{Simenstad1978} and five ``generalized schemes'' 
  % \citep{Nybakken1982,Percival1929,Swan1961,Landry1977,Petipa1979,Harrison1963}, as 
  % we wished to measure
  % directly observed rather than inferred interactions. We further excluded a food web aggregated from several sites in 
  % the Pacific Ocean which were widely spread in both latitude and longitude and at markedly different successional stages
  % \citep{Vinogradov1978}, such that it is not clear which site the food web represents. 
  % This left 223 food webs with widely varying levels of taxonomic resolution, 11 of which contained humans.

@@

  In general, link density, generality, and vulnerability decreased as the
  proportion of basal resources in a web increased (Fig.~\ref{B}). For
  estuarine, lake, and stream food webs these decreases were more gradual towards the
  poles ($\beta_{Latitude}$=-0.007, $p$=0.006 for link density;
  $\beta_{Latitude}$=-0.009, $p$=0.002 for generality; and
  $\beta_{Latitude}$=-0.007, $p$=0.006 for vulnerability).
  In marine food webs there was very little effect of latitude on
  the strength of scaling ($\beta_{Latitude:Marine}$=0.009, $p$=0.047; 
  $\beta_{Latitude:Marine}$=0.011, $p$=0.028; 
  and $\beta_{Latitude:Marine}$=0.009, $p$=0.047 respectively),
  while in terrestrial food webs link density, generality, and vulnerability
  decreased more gradually towards the 
  equator ($\beta_{Latitude:Terrestrial}$=0.013, $p$=0.002; 
  $\beta_{Latitude:Terrestrial}$=0.015, $p$=0.001; 
  and $\beta_{Latitude:Terrestrial}$=0.013, $p$=0.002).

@@

  The lack of a latitudinal gradient in species richness in this dataset contrasts
  strongly with other studies~\citep{Schemske2009a,Macpherson2002,Kaufman1995}. As numbers of species and links included in a food web
  vary strongly with sampling effort as well as with the underlying diversity of the study
  area, it is possible that the lack of latitudinal trends here is a result of researchers
  tending to expend similar amounts of sampling effort across studies. This could result in
  food webs describing species-rich tropical communities omitting more species and links
  than species-poor arctic communities. In addition, it is worth noting that gradients in
  species richness are generally measured for a single taxonomic group at a time (e.g.~\citep{Kaufman1995}).
  It is possible that these taxa are not well-represented in our food webs and that the
  dominant taxa in them do not have an underlying latitudinal gradient in richness. In
  either case, the lack of association between species richness
  and latitude in any ecosystem type nevertheless means that any effect of latitude on scaling relationships
  between species richness and other properties is not being driven by underlying variation in
  species richness, allowing us to more clearly assess effects of latitude on scaling with 
  species richness and proportions of species in different trophic levels.

  Scaling of links per species, generality, and vulnerability with species
  richness varied strongly across ecosystem types. In estuarine, marine, and
  terrestrial food webs scaling of each property varied little with latitude.
  This is consistent with the idea that species' niche breadths do not vary
  systematically with temperature and productivity but that the niche space
  might be larger in species-rich communities~\citep{Davies2007}. Rather than
  niche space depending on temperature and productivity, it may be that species
  diversity itself affects the niche space available to species (although climate
  may affect speciation rates and therefore the diversity in a region~\citep{Currie2004}).

  Unlike other ecosystem types, in lake and stream food webs 
  the scaling of generality was stronger in higher-latitude food webs. 
  In lake food webs, this trend was echoed in the scaling relationships
  between species richness and links per species and vulnerability.
  This means that species in tropical freshwater
  communities gain fewer additional feeding links per additional species in the
  web and that species in tropical lakes also gain fewer predators, and fewer links
  in general, per additional species. These trends are consistent with the hypothesis
  that greater stability in the tropics leads to narrower niches~\citep{Brown2004} and a higher proportion of
  specialists. This may be partly due to high-latitude species
  tending to switch between different seasonally-available
  prey~\citep{Isaac2012,Wilhelm1999} while tropical freshwater ecosystems may have more stable composition.

@@

  Our results were consistent with the latitude-niche breadth hypothesis in estuarine, marine,
  and terrestrial communities but instead consistent with the hypothesis of greater specialization
  in the tropics in stream and lake food webs. This suggests that different mechanisms may structure
  food webs in different habitat types and that freshwater food webs in particular may be strongly
  affected by seasonal variation. In addition, different relationships between latitude and niche
  breadth in different habitat types goes some way towards explaining the equivocal support for
  the opposing hypotheses of~\citet{Brown2004} and ~\citet{Davies2007}; both have merit but appear
  to apply to different systems.

@@

% \section*{Tables}

% \begin{table}[!h]
% \caption{Intercepts and slopes for the best-fitting model describing the relationship between species-richness, mean links per species, mean generality, and mean vulnerability and latitude. Note that none
% of the best-fitting models included a term for latitude or an interaction between latitude and ecosystem type. Each `NA' indicates a term that was not included in the corresponding best-fit model.}
% \label{Latlms}
% \begin{tabular}{l | l l  l l  l l  l l}

% \hline
% \multirow{2}{*}{Property} & \multicolumn{2}{|c}{Intercept} & \multicolumn{2}{|c}{Stream} & \multicolumn{2}{|c}{Lake} & \multicolumn{2}{|c}{Terrestrial}\\
% & $\alpha_{0}$ & $p$-value & $\alpha_{2}$ & $p$-value & $\alpha_{2}$ & $p$-value & $\alpha_{2}$ &$p$-value\\
% \hline
% Species richness  & 34.6 & \textless0.001 & 22.4 & 0.167 \textless0.001 & NA & NA & NA & NA \\
% Links per species & 3.53 & \textless0.001 & NA & NA & NA & NA & NA & NA \\
% Generality        & 7.32 & \textless0.001 & NA & NA & -1.98 & 0.044 & -3.77 & \textless0.001 \\
% Vulnerability     & 3.53 & \textless0.001 & NA & NA & NA & NA & NA & NA \\
% \hline
% \end{tabular}
% \end{table}

% \begin{table}[!h]
% \caption{Intercepts and slopes for the best-fitting model describing the relationship between species-richness, mean links per species, mean generality, and mean vulnerability and latitude in trophic-species versions of food webs. Note that none
% of the best-fitting models included a term for latitude or an interaction between latitude and ecosystem type. `NA' a term that was not included in the corresponding best-fit model.}
% \label{Latlms}
% \begin{tabular}{l | l l  l l  l l  l l}

% \hline
% \multirow{2}{*}{Property} & \multicolumn{2}{|c}{Intercept} & \multicolumn{2}{|c}{Stream} & \multicolumn{2}{|c}{Lake} & \multicolumn{2}{|c}{Terrestrial}\\
% & $\alpha_{0}$ & $p$-value & $\alpha_{2}$ & $p$-value & $\alpha_{2}$ & $p$-value & $\alpha_{2}$ &$p$-value\\
% \hline
% Species richness  & 34.6 & \textless0.001 & 22.4 & 0.167 \textless0.001 & NA & NA & NA & NA \\
% Links per species & 3.53 & \textless0.001 & NA & NA & NA & NA & NA & NA \\
% Generality        & 7.32 & \textless0.001 & NA & NA & -1.98 & 0.044 & -3.77 & \textless0.001 \\
% Vulnerability     & 3.53 & \textless0.001 & NA & NA & NA & NA & NA & NA \\
% \hline
% \end{tabular}
% \end{table}

% \begin{table}[!h]
% \caption{Intercepts and slopes for best-fit models describing the effect of latitude on the scaling of mean links per species, mean generality, and mean vulnerability with species richness. Each `NA' indicates a term that was not included in the corresponding best-fit model. These models refer to the original versions of food webs. }
% \label{Bestfits}
% \begin{tabular}{l | l l  l l  l l }
% Effect          & \multicolumn{2}{c|}{Links per species} & \multicolumn{2}{c|}{Generality} 
%                 & \multicolumn{2}{c}{Vulnerability}\\
%                 & $\beta$ & $p$-value     &  $\beta$ & $p$-value    &  $\beta$ & $p$-value  \\
% \hline
% Intercept               &-0.453 & \textless0.001 & -0.249 & \textless0.001 & -0.453 & \textless0.001 \\
% log(species)            & 0.419 & \textless0.001 &  0.663 & \textless0.001 &  0.419 & \textless0.001 \\
% log(species):Lake       & 0.119 & \textless0.001 & -0.185 & 0.016          &  0.119 & \textless0.001 \\
% log(species):Marine     & 0.255 & \textless0.001 &  0.090 & 0.002          &  0.255 & \textless0.001 \\
% log(species):Terrestrial& 0.207 & 0.002          & -0.122 & \textless0.001 &  0.207 & 0.002 \\
% log(species):Latitude   & 0.004 & \textless0.001 & -0.001 & 0.199          &  0.003 & \textless0.001 \\
% log(species):Latitude:Marine&-0.004 & 0.007      &  NA    &  NA            & -0.005 & 0.007 \\
% log(species):Latitude:Terrestrial&-0.005 & 0.002 &  NA    &  NA            & -0.005 & 0.002 \\
% log(species):Latitude:Lake & NA & NA             &  0.006 & 0.002          &  NA    & NA    \\
% \hline
% $R^{2} (adjusted)$   & \multicolumn{2}{c|}{0.706} & \multicolumn{2}{c|}{0.658} & \multicolumn{2}{c}{0.706} \\
% \hline
% \end{tabular}
% \end{table}

% \begin{table}[!h]
% \caption{Intercepts and slopes for best-fit models describing the effect of latitude on the scaling of mean links per species, mean generality, and mean vulnerability with species richness. Each `NA' indicates a term that was not included in the corresponding best-fit model. These models refer to the trophic species versions of food webs. }
% \label{Bestfits}
% \begin{tabular}{l | l l  l l  l l }
% Effect          & \multicolumn{2}{c|}{Links per species} & \multicolumn{2}{c|}{Generality} 
%                 & \multicolumn{2}{c}{Vulnerability}\\
%                 & $\beta$ & $p$-value     &  $\beta$ & $p$-value    &  $\beta$ & $p$-value  \\
% \hline
% Intercept               & -0.561 & \textless0.001 & -0.299 & \textless0.001 & -0.561 & \textless0.001 \\
% log(species)            &  0.733 & \textless0.001 &  0.622 & \textless0.001 &  0.733 & \textless0.001 \\
% log(species):Lake       & -0.118 & 0.047          & -0.115 & 0.078          & -0.118 & 0.047 \\
% log(species):Marine     &  0.051 & 0.040          &  0.120 & \textless0.001 &  0.051 & 0.040 \\
% log(species):Stream     & -0.161 & 0.014          & -0.057 & 0.421          & -0.161 & 0.014 \\
% log(species):Latitude   & -0.001 & 0.081          & -0.001 & 0.273          & -0.001 & 0.081 \\
% log(species):Lat:Lake   &  0.006 & \textless0.001 &  0.006 & \textless0.001 &  0.006 & \textless0.001 \\
% log(species):Lat:Stream &  0.004 & 0.018          &  0.003 & 0.059          &  0.004 & 0.018 \\
% \hline
% $R^{2} (adjusted)$   & \multicolumn{2}{c|}{0.706} & \multicolumn{2}{c|}{0.658} & \multicolumn{2}{c}{0.706} \\
% \hline
% \end{tabular}
% \end{table}

>>>>

\title{Generality in food webs scales with species richness but not with latitude}
\author{Alyssa R. Cirtwill$^{1,2}$, Daniel B. Stouffer$^{1}$, Tamara N. Romanuk$^{2}$}
\date{$^1$School of Biological Sciences\\University of Canterbury\\
Private Bag 4800\\Christchurch 8140, New Zealand}
\medskip$^2$Department of Biology\\Life Science Centre\\
Dalhousie University\\1355 Oxford St.\\P0 BOX 15000\\
Halifax, NS B3H 4R2\\Canada\\}

@@

Several properties of food webs---the networks of feeding links between
species---vary systematically with the species richness of the underlying
community.  Under the ``latitude-niche breadth hypothesis'', which predicts
that species in the tropics will tend to evolve narrower niches, one might
expect that these scaling relationships could also be affected by latitude. To
test this hypothesis, we analysed the scaling relationships between species
richness and average generality, vulnerability, and links per species across a
set of 163 empirical food webs.  We also investigated scaling relationships
between the three food-web properties and the proportions of the web made up
by basal resources, intermediate consumers, and top predators. While we
observed no effect of latitude on scaling relationships in the estuarine,
marine, and terrestrial food webs, there were strong effects of latitude on
scaling relationships in the freshwater food webs. In these communities, the
latitude-niche breadth hypothesis appears to hold true while in other habitat
types niches appear to be broader in the tropics. These contrasting findings
indicate that it may be more important to account for habitat than latitude
when exploring gradients in food-web structure.

@@

@@

    In general, link density, generality, and vulnerability decreased as the
    proportion of basal resources in a web increased (Fig.~\ref{B}). For
    estuarine, lake, and stream food webs these decreases were more gradual towards the
    poles ($\beta_{Latitude}$=-0.007, $p$=0.006 for link density;
    $\beta_{Latitude}$=-0.009, $p$=0.002 for generality; and
    $\beta_{Latitude}$=-0.007, $p$=0.006 for vulnerability).
    In marine food webs there was very little effect of latitude on
    the strength of scaling ($\beta_{Latitude:Marine}$=0.009, $p$=0.047; 
    $\beta_{Latitude:Marine}$=0.011, $p$=0.028; 
    and $\beta_{Latitude:Marine}$=0.009, $p$=0.047 respectively),
    while in terrestrial food webs link density, generality, and vulnerability each
    decreased more gradually towards the 
    equator ($\beta_{Latitude:Terrestrial}$=0.013, $p$=0.002; 
    $\beta_{Latitude:Terrestrial}$=0.015, $p$=0.001; 
    and $\beta_{Latitude:Terrestrial}$=0.013, $p$=0.002).

@@

  The lack of a latitudinal gradient in species richness in this dataset contrasts
  strongly with other studies~\citep{Schemske2009a,Macpherson2002,Kaufman1995}. As numbers of species and links included in a food web
  vary strongly with sampling effort as well as with the underlying diversity of the study
  area, it is possible that the lack of latitudinal trends here is a result of researchers
  tending to expend similar amounts of sampling effort across studies. This could result in
  food webs describing species-rich tropical communities omitting more species and links
  than species-poor arctic communities. In addition, it is worth noting that gradients in
  species richness are generally measured for a single taxonomic group at a time (e.g.~\citep{Kaufman1995}).
  It is possible that these taxa are not well-represented in our food webs and that the
  dominant taxa in them do not have an underlying latitudinal gradient in richness. 
  In either case, the lack of a strong association between
  species richness and latitude in any ecosystem type
  means that any effect of latitude on other scaling 
  relationships is not being driven by an underlying 
  latitudinal gradient in species richness.
  The lack of confounding effects of latitude allows us to 
  more clearly assess effects of latitude on scaling with 
  species richness and proportions of species in different 
  trophic levels.

  Scaling of links per species, generality, and vulnerability with species
  richness varied strongly across ecosystem types. In estuarine, marine, and
  terrestrial food webs scaling of each property varied little with latitude.
  This is consistent with the idea that species' niche breadths do not vary
  systematically with temperature and productivity but that the niche space
  might be larger in species-rich communities~\citep{Davies2007}. Rather than
  niche space depending on temperature and productivity, it may be that species
  diversity itself affects the niche space available to species (although climate
  may affect speciation rates and therefore the diversity in a region~\citep{Currie2004}). 
  For example, as plant diversity increases both the 
  variety of food available to herbivores and the structural variety of the habitat will also increase.

  Unlike other ecosystem types, the scaling of generality
  in lake and stream food webs was stronger (i.e., 
  increased more steeply) in higher-latitude food webs. 
  In lake food webs, this trend was echoed in the scaling relationships
  between species richness and links per species and vulnerability.
  This means that species in tropical freshwater
  communities gain fewer additional feeding links per additional species in the
  web and that species in tropical lakes also gain fewer predators, and fewer links
  in general, per additional species. These trends are consistent with the hypothesis
  that greater stability in the tropics leads to narrower niches~\citep{Vazquez2004} and a higher proportion of
  specialists. This may be partly due to high-latitude species
  tending to switch between different seasonally-available
  prey~\citep{Isaac2012,Wilhelm1999} while tropical freshwater ecosystems may have more stable composition.

@@

  Our results were inconsistent with the latitude-niche breadth hypothesis in estuarine, marine,
  and terrestrial communities but consistent with the hypothesis of greater specialization
  in the tropics in stream and lake food webs. This suggests that different mechanisms may structure
  food webs in different habitat types and that freshwater food webs in particular may be strongly
  affected by seasonal variation. In addition, different relationships between latitude and niche
  breadth in different habitat types goes some way towards explaining the equivocal support for
  the opposing hypotheses of narrower niches in the tropics~\citep{Vazquez2004} and broader 
  niche space in the tropics~\citep{Davies2007}; both have merit but appear
  to apply to different systems.

@@
